---
title: "The Glass Ceiling: Does Education Equitably Predict Wages for Ontario Immigrants?"
author:
  - "Moiz Mohammad"
  - "Asad Mirza"
  - "Ishan Amin"
date: "`r Sys.Date()`"
documentclass: article
fontsize: 11pt
geometry: margin=0.5in
header-includes:
  - '\linespread{1.15}\selectfont'
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  eval = FALSE
)
ttest_to_table <- function(test_obj, group_1, group_2, diff_label) {
  estimate_names <- names(test_obj$estimate)

  get_group_mean <- function(group_label, fallback_index) {
    idx <- which(grepl(group_label, estimate_names, fixed = TRUE))
    if (length(idx) == 1) {
      return(unname(test_obj$estimate[[idx]]))
    }
    unname(test_obj$estimate[[fallback_index]])
  }

  mean_group_1 <- get_group_mean(group_1, 1)
  mean_group_2 <- get_group_mean(group_2, 2)

  tibble::tibble(
    comparison = paste(group_1, "vs", group_2),
    mean_group_1 = mean_group_1,
    mean_group_2 = mean_group_2,
    mean_difference = mean_group_1 - mean_group_2,
    t_statistic = unname(test_obj$statistic),
    df = unname(test_obj$parameter),
    p_value = test_obj$p.value,
    ci_lower = test_obj$conf.int[1],
    ci_upper = test_obj$conf.int[2],
    difference_definition = diff_label
  )
}

ttest_to_long_table <- function(test_obj, group_1, group_2, diff_label) {
  summary_tbl <- ttest_to_table(test_obj, group_1, group_2, diff_label)
  tibble::tibble(
    Metric = c(
      "Comparison",
      paste("Mean:", group_1),
      paste("Mean:", group_2),
      "Mean difference",
      "Difference definition",
      "t-statistic",
      "Degrees of freedom",
      "p-value",
      "95% CI lower",
      "95% CI upper"
    ),
    Value = c(
      summary_tbl$comparison,
      sprintf("%.3f", summary_tbl$mean_group_1),
      sprintf("%.3f", summary_tbl$mean_group_2),
      sprintf("%.3f", summary_tbl$mean_difference),
      summary_tbl$difference_definition,
      sprintf("%.3f", summary_tbl$t_statistic),
      sprintf("%.3f", summary_tbl$df),
      format(summary_tbl$p_value, digits = 3, scientific = TRUE),
      sprintf("%.3f", summary_tbl$ci_lower),
      sprintf("%.3f", summary_tbl$ci_upper)
    )
  )
}

run_immigrant_status_test <- function(data, mean_caption, test_caption) {
  groups <- c("Born in Canada", "Established immigrants, 10+ years")

  mean_tbl <- stats::aggregate(
    wage_both ~ immigrant_status,
    data = data,
    FUN = function(x) mean(x, na.rm = TRUE)
  )
  names(mean_tbl)[2] <- "mean_wage"
  mean_tbl <- tibble::as_tibble(mean_tbl)

  test_data <- data[data$immigrant_status %in% groups, , drop = FALSE]

  test_obj <- stats::t.test(wage_both ~ immigrant_status, data = test_data)

  print(knitr::kable(mean_tbl, caption = mean_caption))

  knitr::kable(
    ttest_to_long_table(
      test_obj,
      groups[1],
      groups[2],
      paste(groups, collapse = " - ")
    ),
    caption = test_caption
  )
}
bootstrap_mean_diff <- function(group_a, group_b, reps = 1000) {
  group_a <- group_a[!is.na(group_a)]
  group_b <- group_b[!is.na(group_b)]

  replicate(reps, {
    mean(sample(group_a, size = length(group_a), replace = TRUE)) -
      mean(sample(group_b, size = length(group_b), replace = TRUE))
  })
}

bootstrap_from_groups <- function(data, group_col, group_a, group_b, reps = 1000) {
  vec_a <- data$wage_both[data[[group_col]] == group_a]
  vec_b <- data$wage_both[data[[group_col]] == group_b]

  bootstrap_mean_diff(vec_a, vec_b, reps = reps)
}
```
\newpage
# Project overview

This project examines whether **education** predicts **wages** in a similar way for **immigrants in Ontario** as it might for other groups, or whether systematic differences suggest barriers consistent with a **glass ceiling** (e.g., credentials not translating equally into earnings). We combine exploratory work, inference, and modeling to connect Ontario labour data to that question.

# Research questions

1. How strongly is educational attainment associated with wages among Ontario residents with an immigrant background?
2. Does the education--wage gradient differ by gender, region, immigrant class, or years since landing?
3. After controlling for relevant covariates, do disparities remain that suggest education does **not** equitably predict wages for this population?

```{r libraries, eval=TRUE}
# Shared packages;
library(tidyverse)
library(knitr)
library(kableExtra)
```

```{r data-load, eval=TRUE}
wages_raw <- readr::read_csv(
  "data.csv",
  trim_ws = TRUE,
  show_col_types = FALSE
)
wages <- wages_raw
```

# The Foundation and Visuals

## Background: Ontario Ministry of Labour, Immigration, Training and Skills Development

The Ontario Ministry of Labour, Immigration, Training and Skills Development (MLITSD) publishes the Wages by Education Level dataset (2006–2024) to monitor labour market equity and inform evidence-based policy. This analysis utilizes that data to evaluate how educational credentials translate into wage returns, specifically focusing on the "glass ceiling" often encountered by immigrants.

Despite Ontario’s reliance on high-skilled immigration through programs like the OINP, a persistent wage gap exists between Canadian-born workers and newcomers. By examining median hourly wages disaggregated by status, age, and education, this study acts as a diagnostic tool to identify where the education-wage relationship breaks down. The following sections define our variables and utilize exploratory visualizations to quantify how equitably credentials are rewarded in Ontario’s modern economy.

## Data Dictionary

The dataset is organized with one row per unique combination of year, geography, immigrant status, education level, and age group. The primary metric is the median hourly wage (CAD), disaggregated into three columns: Both sexes, Men, and Women. This analysis utilizes the Both sexes aggregate as the default outcome variable.

| Column name  | Description                                                                                      | Type / Units                                       | Notes                                                                                                               |
|--------------------------|--------------------------------------------------------------------------------------------------|----------------------------------------------------|---------------------------------------------------------------|
| `YEAR`           | Calendar year of the observation                                                                 | Integer (2006–2024)                                | No missing values                                                                                                   |
| `GEOGRAPHY`      | Geographic scope of the record                                                                   | Categorical (Canada / Ontario)                     | Filtered to Ontario only for this analysis                                                                          |
| `IMMIGRANT`      | Immigration classification of the worker group                                                   | Categorical (character)                            | Includes Born in Canada, Non-landed, and landed immigrant sub-groups; rows containing "Total" dropped               |
| `TYPE OF WORK`   | Employment classification                                                                        | Categorical (All employees / Full-time / Part-time)| Filtered to "All employees" for EDA; full-time-only is a useful sensitivity filter                                  |
| `WAGE RATE`      | Descriptor of the wage measure recorded in the file                                              | Categorical — single value: "Median hourly wage"  | Not used analytically; only one value exists in the dataset                                                         |
| `EDUCATION`      | Highest level of schooling attained by the worker group                                          | Ordered factor (8 levels)                          | Leading whitespace trimmed; ordered from lowest (0–8 years) to highest (Above bachelor's degree); "Total" rows dropped |
| `AGE GROUP`      | Age cohort of the worker group                                                                   | Categorical (15+, 25+, 25–34, 25–54, 25–64)       | Filtered to "25+" for main analysis to focus on prime working-age adults                                            |
| `Both sexes`     | Median hourly wage (CAD/hr) for both sexes combined — **primary outcome variable**               | Numeric (CAD/hour)                                 | No missing values in Ontario subset                                                                                  |
| `Men`            | Median hourly wage (CAD/hr) for men                                                              | Numeric (CAD/hour)                                 | Retained for gender-disaggregated analysis in later sections                                                        |
| `Women`          | Median hourly wage (CAD/hr) for women                                                            | Numeric (CAD/hour)                                 | Retained for gender-disaggregated analysis in later sections                                                        |


## Data Cleaning

Rows where `EDUCATION` or `IMMIGRANT` contained the word `"Total"` were removed prior to analysis to prevent double-counting, as these aggregate rows represent summaries across sub-groups already present in the data. The dataset was further filtered to Ontario only, since Canada-level rows would conflate Ontario's labour market with national trends outside this project's scope.

```{r data-clean, eval=TRUE}
library(tidyverse)

# Load the raw dataset
wages_raw <- readr::read_csv(
  "data.csv",
  trim_ws        = TRUE,
  show_col_types = FALSE
)
wages <- wages_raw %>%
  # Standardize column names to snake_case
  rename(
    year             = YEAR,
    geography        = GEOGRAPHY,
    immigrant_status = IMMIGRANT,
    work_type        = `TYPE OF WORK`,
    wage_type        = `WAGE RATE`,
    education_level  = EDUCATION,
    age_group        = `AGE GROUP`,
    wage_both        = `Both sexes`,
    wage_men         = Men,
    wage_women       = Women
  ) %>%
  # Focus on Ontario only (drop Canada-level aggregate rows)
  filter(str_trim(geography) == "Ontario") %>%
  filter(
    !str_detect(education_level, regex("total", ignore_case = TRUE)),
    !str_detect(immigrant_status, regex("total", ignore_case = TRUE))
  ) %>%
  mutate(across(where(is.character), str_trim)) %>%
  # Reorder education_level as an ordered factor from lowest to highest attainment
  mutate(
    education_level = factor(
      education_level,
      levels = c(
        "0 - 8 years",
        "Some high school",
        "High school graduate",
        "Some post-secondary",
        "Post-secondary certificate or diploma",
        "University degree",
        "Bachelor's degree",
        "Above bachelor's degree"
      ),
      ordered = TRUE
    )
  ) %>%
  filter(age_group == "25 +", work_type == "All employees")
glimpse(wages)
```

---

## Exploratory Data Analysis

The table below presents the mean, median, and standard deviation of the median hourly wage (`wage_both`, both sexes) for each education level, pooled across all years and immigrant status categories in the Ontario, age 25+ subset.

```{r table1-summary, eval=TRUE}
wages %>%
  group_by(education_level) %>%
  summarise(
    n           = n(),
    mean_wage   = mean(wage_both, na.rm = TRUE),
    median_wage = median(wage_both, na.rm = TRUE),
    sd_wage     = sd(wage_both, na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  knitr::kable(
    digits = 2,
    col.names = c(
      "Education Level", "N (obs.)",
      "Mean Hourly Wage ($)", "Median Hourly Wage ($)", "Std. Dev. ($)"
    ),
    format = "latex"
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width    = FALSE,
    font_size     = 10
  )
```

Table 1 reveals a clear positive wage gradient: mean hourly wages increase at each successive level of educational attainment, with university-level workers earning substantially more than those with eight or fewer years of schooling.

```{r graph1-immigrant-trend, eval=TRUE, fig.width=6, fig.height=4, fig.align='center', out.width='75%'}
wages_immig <- wages %>%
  mutate(
    immig_group = case_when(
      str_detect(immigrant_status, regex("born in canada", ignore_case = TRUE)) ~ "Born in Canada",
      str_detect(immigrant_status, regex("landed", ignore_case = TRUE)) ~ "Total Landed Immigrants",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(immig_group)) %>%
  group_by(year, immig_group) %>%
  summarise(mean_wage = mean(wage_both, na.rm = TRUE), .groups = "drop")

ggplot(wages_immig, aes(
  x = year, y = mean_wage,
  colour = immig_group,
  group = immig_group
)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8, alpha = 0.9) +
  scale_colour_manual(
    values = c("Born in Canada" = "#2a9d8f", "Total Landed Immigrants" = "#f4a261"),
    name   = "Worker Group"
  ) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$", suffix = "/hr")) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 7)) +
  labs(
    title    = "Wage Trends Over Time: Immigrants vs. Canadian-Born Workers",
    subtitle = "Mean of median hourly wages (CAD/hr) — Ontario, age 25+",
    x        = "Year",
    y        = "Mean Wage (CAD/hr)",
    caption  = "Source: Ontario MLITSD. 'Total Landed Immigrants' includes all arrival categories."
  ) +
  theme_minimal(base_size = 9) +
  theme(
    axis.title.y = element_text(margin = margin(r = 10)),
    plot.title = element_text(face = "bold", size = 10, hjust = 0.5),
    plot.subtitle = element_text(colour = "grey40", size = 8, hjust = 0.5),
    plot.caption = element_text(colour = "grey50", size = 7, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )
```

```{r graph2-university-immigrant, eval=TRUE, fig.width=6, fig.height=4, fig.align='center', out.width='75%'}
wages_univ <- wages %>%
  filter(education_level == "University degree") %>%
  mutate(
    immig_group = case_when(
      str_detect(immigrant_status, regex("born in canada", ignore_case = TRUE)) ~ "Born in Canada",
      str_detect(immigrant_status, regex("landed", ignore_case = TRUE)) ~ "Total Landed Immigrants",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(immig_group)) %>%
  group_by(year, immig_group) %>%
  summarise(mean_wage = mean(wage_both, na.rm = TRUE), .groups = "drop")

ggplot(wages_univ, aes(
  x = year, y = mean_wage,
  colour = immig_group,
  group = immig_group
)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8, alpha = 0.9) +
  scale_colour_manual(
    values = c("Born in Canada" = "#2a9d8f", "Total Landed Immigrants" = "#f4a261"),
    name   = "Worker Group"
  ) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$", suffix = "/hr")) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 7)) +
  labs(
    title    = "University Degree Holders: Immigrants vs. Canadian-Born",
    subtitle = "Mean of median hourly wages (CAD/hr) — Ontario, age 25+",
    x        = "Year",
    y        = "Mean Wage (CAD/hr)",
    caption  = "Source: Ontario MLITSD. Data filtered to 'University degree' only."
  ) +
  theme_minimal(base_size = 9) +
  theme(
    axis.title.y = element_text(margin = margin(r = 10)),
    plot.title = element_text(face = "bold", size = 10, hjust = 0.5),
    plot.subtitle = element_text(colour = "grey40", size = 8, hjust = 0.5),
    plot.caption = element_text(colour = "grey50", size = 7, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )
```

**Interpretation.** Graphs 1 and 2 illustrate a persistent wage gap between landed immigrants and Canadian-born workers in Ontario (2006–2024) that education alone cannot bridge. Graph 1 establishes that immigrants consistently earn less even when all education levels are pooled, proving the disadvantage is not merely a credentialing issue. Graph 2 isolates university graduates to provide direct evidence of a "glass ceiling"; the fact that this gap has not narrowed over two decades suggests that structural barriers—such as employer discrimination or non-recognition of foreign credentials outweigh temporary integration hurdles like language. This sustained disparity directly motivates the formal hypothesis tests and regression models in the sections that follow.

# Bootstrapping and Hypothesis Testing

## Hypothesis testing

We use Welch two-sample t-tests to assess whether wages differ by immigrant status for "Born in Canada" versus "Established immigrants, 10+ years." The comparison is run in two samples: the overall dataset and the university-only subgroup. This lets us test whether the gap persists even when education is held constant.
Null hypothesis: mean wages are equal between the two groups.
Alternative hypothesis: mean wages differ.

```{r hypothesis-tests-immigrant, eval=TRUE}
immigrant_data <- wages %>%
  filter(immigrant_status %in% c("Born in Canada", "Established immigrants, 10+ years"))
immigrant_test <- t.test(wage_both ~ immigrant_status, data = immigrant_data)
ttest_to_table(
  immigrant_test,
  "Born in Canada",
  "Established immigrants, 10+ years",
  "Born in Canada - Established immigrants, 10+ years"
) %>%
  transmute(
    Sample = "Overall",
    `Mean difference` = round(mean_difference, 3),
    `95% CI` = sprintf("[%.3f, %.3f]", ci_lower, ci_upper),
    `p-value` = format(p_value, digits = 3, scientific = TRUE)
  ) %>%
  kable(caption = "Welch t-test summary: overall sample")
```

```{r hypothesis-tests-university-immigrant, eval=TRUE}
university_data <- wages %>%
  filter(
    education_level == "University degree",
    immigrant_status %in% c("Born in Canada", "Established immigrants, 10+ years")
  )
university_immigrant_test <- t.test(wage_both ~ immigrant_status, data = university_data)
ttest_to_table(
  university_immigrant_test,
  "Born in Canada",
  "Established immigrants, 10+ years",
  "Born in Canada - Established immigrants, 10+ years"
) %>%
  transmute(
    Sample = "University only",
    `Mean difference` = round(mean_difference, 3),
    `95% CI` = sprintf("[%.3f, %.3f]", ci_lower, ci_upper),
    `p-value` = format(p_value, digits = 3, scientific = TRUE)
  ) %>%
  kable(caption = "Welch t-test summary: university-only sample")
```

**Interpretation.** In both samples, the estimated mean difference (Born in Canada - Established immigrants) is positive, and each 95% confidence interval lies entirely above 0. The p value for each is also incredibly small(lower than 0.05). We therefore reject the null hypothesis in both tests. The overall-sample result indicates a broad immigrant-status wage gap, while the university-only result shows the gap remains even among workers with the same higher-education category. Together, these tests support the report's core claim that wage returns are not distributed equally across immigrant groups.

## Bootstrapping

Bootstrapping provides a non-parametric check of the immigrant-status wage gaps by repeatedly resampling the observed data with replacement. Unlike the t-test, it does not rely as strongly on normality assumptions for the sampling distribution of the mean difference. Here, we bootstrap the same two comparisons from the hypothesis tests (overall and university-only) to verify that the direction and size of the wage gap are stable under resampling.

```{r bootstrap, fig.cap="Bootstrap distributions of mean wage differences for the two immigrant-status comparisons.", eval=TRUE, fig.width=7, fig.height=4.2, fig.align='center', out.width='88%', fig.pos='H'}
set.seed(123)
bootstrap_specs <- tibble(
  comparison_plot = c(
    "Born in Canada - Established immigrants, 10+ years",
    "Born in Canada - Established immigrants, 10+ years\n(University only)"
  ),
  comparison_table = c(
    "Born in Canada - Established immigrants, 10+ years",
    "Born in Canada - Established immigrants, 10+ years (University only)"
  ),
  group_var = c("immigrant_status", "immigrant_status"),
  group_a = c("Born in Canada", "Born in Canada"),
  group_b = c("Established immigrants, 10+ years", "Established immigrants, 10+ years"),
  subset_education = c(NA_character_, "University degree")
)
bootstrap_results <- bootstrap_specs %>%
  mutate(
    boot = pmap(
      list(group_var, group_a, group_b, subset_education),
      \(group_var, group_a, group_b, subset_education) {
        data_use <- wages
        if (!is.na(subset_education)) {
          data_use <- data_use %>% filter(education_level == subset_education)
        }
        bootstrap_from_groups(data_use, group_var, group_a, group_b, reps = 1000)
      }
    ),
    bootstrap_mean_difference = map_dbl(boot, mean),
    ci_lower = map_dbl(boot, ~ quantile(.x, 0.025)),
    ci_upper = map_dbl(boot, ~ quantile(.x, 0.975))
  )
bootstrap_plot_data <- bootstrap_results %>%
  select(comparison_plot, boot) %>%
  unnest_longer(boot, values_to = "mean_difference") %>%
  mutate(comparison = factor(comparison_plot, levels = bootstrap_specs$comparison_plot))
bootstrap_ci_lines <- bootstrap_results %>%
  select(comparison_plot, ci_lower, ci_upper) %>%
  pivot_longer(c(ci_lower, ci_upper), values_to = "xintercept") %>%
  mutate(comparison = factor(comparison_plot, levels = bootstrap_specs$comparison_plot))
ggplot(bootstrap_plot_data, aes(x = mean_difference)) +
  geom_histogram(bins = 30, fill = "#8fb9a8", color = "white") +
  geom_vline(
    data = bootstrap_ci_lines,
    aes(xintercept = xintercept),
    color = "#c0392b",
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  facet_wrap(~comparison, ncol = 2, scales = "free_y") +
  labs(
    title = "Bootstrap Mean Difference Distributions",
    x = "Bootstrap mean difference",
    y = "Count"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
bootstrap_summary <- bootstrap_results %>%
  transmute(
    Comparison = comparison_table,
    `Bootstrap mean diff` = bootstrap_mean_difference,
    `95% CI` = paste0("[", round(ci_lower, 3), ", ", round(ci_upper, 3), "]")
  )
kable(
  bootstrap_summary,
  digits = 3,
  caption = "Bootstrap 95\\% confidence intervals for mean wage differences"
) %>%
  kableExtra::kable_styling(
    latex_options = "hold_position",
    full_width = FALSE,
    font_size = 9
  )
```

**Interpretation.** Both bootstrap intervals stay above 0, matching the t-test conclusions: Canadian-born workers earn more on average than established immigrants, both overall and within university graduates. This indicates the finding is robust to distributional assumptions.

# The Model and Conclusion

## Regression analysis
```{r regression, eval=TRUE}
# Prepare Data
wages_model <- wages %>%
  mutate(
    education_level = relevel(factor(education_level, ordered = FALSE), ref = "High school graduate"),
    immigrant_status = relevel(factor(immigrant_status), ref = "Born in Canada")
  ) %>%
  filter(!is.na(wage_both), !is.na(education_level), !is.na(immigrant_status))
wages_model <- wages_model %>% mutate(year_centered = year - 2006)
# Regression Model
fit_mlr <- lm(wage_both ~ education_level + immigrant_status + year_centered, data = wages_model)
# Extract Results
coef_mat <- summary(fit_mlr)$coefficients
# Create Table with Results
reg_table <- tibble::tibble(term = rownames(coef_mat), estimate = coef_mat[, 1], p_value = coef_mat[, 4])
# Clean labels
reg_table_clean <- reg_table %>%
  mutate(term = stringr::str_remove_all(term, "education_level|immigrant_status")) %>%
  mutate(term = ifelse(term == "(Intercept)", "Baseline (High School / Born in Canada)", term))
# Keep only important variables
important_terms <- c(
  "Baseline (High School / Born in Canada)",
  "0 - 8 years",
  "Some post-secondary",
  "Bachelor's degree",
  "University degree",
  "Above bachelor's degree",
  "Established immigrants, 10+ years",
  "Recent immigrants, 5+ to 10 years",
  "Very recent immigrants, 5 years or less",
  "year_centered"
)
# Name cleaning
reg_table_compact <- reg_table_clean %>%
  filter(term %in% important_terms) %>%
  mutate(
    term = dplyr::case_when(
      term == "Baseline (High School / Born in Canada)" ~ "Reference: High school graduate + Born in Canada (year 2006)",
      term %in% c("0 - 8 years", "Some post-secondary", "Bachelor's degree", "University degree", "Above bachelor's degree") ~
        paste0("Education: ", term, " (vs High school graduate)"),
      term %in% c("Established immigrants, 10+ years", "Recent immigrants, 5+ to 10 years", "Very recent immigrants, 5 years or less") ~
        paste0("Immigrant status: ", term, " (vs Born in Canada)"),
      term == "year_centered" ~ "Year centered at 2006 (per +1 year)",
      TRUE ~ term
    ),
    p_value = ifelse(p_value < 0.001, "<0.001", sprintf("%.3f", p_value))
  ) %>%
  transmute(Variable = term, Estimate = round(estimate, 3), `p-value` = p_value)
knitr::kable(reg_table_compact, booktabs = TRUE, caption = "Key Regression Coefficients for Median Hourly Wage (CAD/hr)")
```
**Interpretation.** The linear regression model shows that education is strongly associated with higher wages, but the wage gap for those with immigrant-status remains. The baseline group (high school graduates born in Canada in 2006) has a predicted wage of 18.595 CAD/hr. Compared to this baseline, those with university-level education have higher wages (University degree +11.745, Bachelor’s +9.196, Above bachelor’s +16.303 CAD/hr, all p < 0.001 so the results are strongly supported). Those with lower education categories are associated with lower wages. The only education category without a clear difference is some post-secondary (-0.147 CAD/hr, p = 0.744, but as p value is high, results are not as accurate).

Those with an immigrant-status earn less than Canadian-born workers, even after controlling education and year (established immigrants -2.895, recent immigrants 5+ to 10 years -7.954, very recent immigrants -10.259 CAD/hr, all p < 0.001 so the results are strongly supported). The positive year effect (+0.508 CAD/hr, p < 0.001) indicates general wage growth over time but not nearly as high as Canadian-born workers. Overall, the model supports the main conclusion that education is rewarded, but the return is not equally distributed across immigrant groups indicating the presence of a glass-ceiling.


## Cross-validation
```{r cross-validation, eval=TRUE}
# 4-fold cross-validation
set.seed(123)
d <- wages %>%
  mutate(
    education_level = relevel(factor(as.character(education_level), ordered = FALSE), ref = "High school graduate"),
    immigrant_status = relevel(factor(immigrant_status), ref = "Born in Canada"),
    year_centered = year - 2006
  ) %>%
  filter(!is.na(wage_both), !is.na(education_level), !is.na(immigrant_status), !is.na(year_centered))
k <- 4
d <- d %>% mutate(group_ind = sample(1:k, size = nrow(d), replace = TRUE))
mse_vec <- numeric(k)
# Perform cross-validation
for (i in 1:k) {
  d.train <- d %>% filter(group_ind != i)
  d.test <- d %>% filter(group_ind == i)
  lm.mod <- lm(wage_both ~ education_level + immigrant_status + year_centered, data = d.train)
  y_hat <- predict(lm.mod, newdata = d.test)
  mse_vec[i] <- mean((d.test$wage_both - y_hat)^2, na.rm = TRUE)
}
# TABLE 1: Detailed Fold Results
cv_results <- tibble::tibble(Fold = paste0("Fold ", 1:k), `MSE ((CAD/hr)^2)` = round(mse_vec, 3))
knitr::kable(cv_results, booktabs = TRUE, caption = "4-Fold Cross-Validation: Mean Squared Error by Fold", escape = FALSE) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
# TABLE 2: Summary Metrics
overall_mse <- mean(mse_vec)
overall_rmse <- sqrt(overall_mse)
summary_metrics <- tibble::tibble(Metric = c("Average MSE", "Average RMSE"), Unit = c("(CAD/hr)^2", "CAD/hr"), Value = round(c(overall_mse, overall_rmse), 3))
knitr::kable(summary_metrics, booktabs = TRUE, caption = "Overall Model Predictive Performance", col.names = c("Metric", "Unit", "Value")) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```
**Interpretation.** We evaluated the regression model performance using a 4-fold cross-validation. In each fold, the model was trained on 3 parts of the data and tested on the remaining part, then MSE was computed on the test fold. MSE values ranged from 9.105 to 14.291, with an average of 11.649. The corresponding average RMSE is 3.413 CAD/hr, meaning predictions are typically off by about $3.4 per hour on unseen data. Overall, the model shows a fairly stable predictive performance across folds.

## Summary and conclusion

Overall, the evidence indicates that education is strongly associated with higher wages in Ontario, but these returns are not equally distributed across immigrant groups. Analysis shows a clear wage difference by education level and an increasing education-related wage gap over time. Inference results and bootstrap confidence intervals consistently show significant differences in wages by both education and immigrant status, including within the university-educated subgroup.

The regression model confirms that higher education categories are associated with higher hourly wages after controlling for time, while several immigrant-status groups remain at a wage disadvantage relative to Canadian-born workers. Cross-validation results also suggest moderate and stable out-of-sample predictive performance, supporting the reliability of the model findings.

Taken together, these results support the glass-ceiling interpretation that education does improve earnings on average, but these earnings are not equally distributed across groups. This suggests the need for stronger educational recognition and improved newcomer wages, to ensure fairer wage returns to education in Ontario regardless of status.

# Appendix: All code

```{r appendix-all-code, ref.label=c("libraries", "data-load", "table1-summary", "graph1-immigrant-trend", "graph2-university-immigrant", "hypothesis-tests-immigrant", "hypothesis-tests-university-immigrant", "bootstrap", "regression", "cross-validation"), echo=TRUE, eval=FALSE}
```

# References

Ministry of Labour, Immigration, training and Skills Development. ontario.ca. (2025, August 12). https://www.ontario.ca/page/ministry-labour-immigration-training-skills-development 

Ontario.ca. (n.d.). Wages by education level. Wages by education level - Dataset - Ontario Data Catalogue. https://data.ontario.ca/dataset/wages-by-education-level 