---
title: "The Glass Ceiling: Does Education Equitably Predict Wages for Ontario Immigrants?"
author:
  - "Moiz Mohammad"
  - "Asad Mirza"
  - "Ishan Amin"
date: "`r Sys.Date()`"
documentclass: article
fontsize: 11pt
geometry: margin=0.5in
header-includes:
  - '\linespread{1.15}\selectfont'
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
# Set eval=TRUE on analysis chunks once your data path and cleaning are ready.
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  eval = FALSE
)

ttest_to_table <- function(test_obj, group_1, group_2, diff_label) {
  tibble::tibble(
    comparison = paste(group_1, "vs", group_2),
    mean_group_1 = unname(test_obj$estimate[[1]]),
    mean_group_2 = unname(test_obj$estimate[[2]]),
    mean_difference = unname(diff(test_obj$estimate)),
    t_statistic = unname(test_obj$statistic),
    df = unname(test_obj$parameter),
    p_value = test_obj$p.value,
    ci_lower = test_obj$conf.int[1],
    ci_upper = test_obj$conf.int[2],
    difference_definition = diff_label
  )
}

ttest_to_long_table <- function(test_obj, group_1, group_2, diff_label) {
  summary_tbl <- ttest_to_table(test_obj, group_1, group_2, diff_label)
  tibble::tibble(
    Metric = c(
      "Comparison",
      paste("Mean:", group_1),
      paste("Mean:", group_2),
      "Mean difference",
      "Difference definition",
      "t-statistic",
      "Degrees of freedom",
      "p-value",
      "95% CI lower",
      "95% CI upper"
    ),
    Value = c(
      summary_tbl$comparison,
      sprintf("%.3f", summary_tbl$mean_group_1),
      sprintf("%.3f", summary_tbl$mean_group_2),
      sprintf("%.3f", summary_tbl$mean_difference),
      summary_tbl$difference_definition,
      sprintf("%.3f", summary_tbl$t_statistic),
      sprintf("%.3f", summary_tbl$df),
      format(summary_tbl$p_value, digits = 3, scientific = TRUE),
      sprintf("%.3f", summary_tbl$ci_lower),
      sprintf("%.3f", summary_tbl$ci_upper)
    )
  )
}
```
\newpage
# Project overview

This project examines whether **education** predicts **wages** in a similar way for **immigrants in Ontario** as it might for other groups, or whether systematic differences suggest barriers consistent with a **glass ceiling** (e.g., credentials not translating equally into earnings). We combine exploratory work, inference, and modeling to connect Ontario labour data to that question.

# Research questions

1. How strongly is educational attainment associated with wages among Ontario residents with an immigrant background?
2. Does the education--wage gradient differ by gender, region, immigrant class, or years since landing?
3. After controlling for relevant covariates, do disparities remain that suggest education does **not** equitably predict wages for this population?

```{r libraries, eval=TRUE}
# Shared packages;
library(tidyverse)
library(knitr)
library(kableExtra) # optional: nicer tables in PDF
```

```{r data-load, eval=TRUE}
wages_raw <- readr::read_csv(
  "data.csv",
  trim_ws = TRUE,
  show_col_types = FALSE
)

wages <- wages_raw
```

# The Foundation and Visuals

## Background: Ontario Ministry of Labour, Immigration, Training and Skills Development

The Ontario Ministry of Labour, Immigration, Training and Skills Development (MLITSD) publishes the Wages by Education Level dataset (2006–2024) to monitor labour market equity and inform evidence-based policy. This analysis utilizes that data to evaluate how educational credentials translate into wage returns, specifically focusing on the "glass ceiling" often encountered by immigrants.

Despite Ontario’s reliance on high-skilled immigration through programs like the OINP, a persistent wage gap exists between Canadian-born workers and newcomers. By examining median hourly wages disaggregated by status, age, and education, this study acts as a diagnostic tool to identify where the education-wage relationship breaks down. The following sections define our variables and utilize exploratory visualizations to quantify how equitably credentials are rewarded in Ontario’s modern economy.

## Data Dictionary

The dataset is organized with one row per unique combination of year, geography, immigrant status, education level, and age group. The primary metric is the median hourly wage (CAD), disaggregated into three columns: Both sexes, Men, and Women. This analysis utilizes the Both sexes aggregate as the default outcome variable.

| Column name  | Description                                                                                      | Type / Units                                       | Notes                                                                                                               |
|--------------------------|--------------------------------------------------------------------------------------------------|----------------------------------------------------|---------------------------------------------------------------|
| `YEAR`           | Calendar year of the observation                                                                 | Integer (2006–2024)                                | No missing values                                                                                                   |
| `GEOGRAPHY`      | Geographic scope of the record                                                                   | Categorical (Canada / Ontario)                     | Filtered to Ontario only for this analysis                                                                          |
| `IMMIGRANT`      | Immigration classification of the worker group                                                   | Categorical (character)                            | Includes Born in Canada, Non-landed, and landed immigrant sub-groups; rows containing "Total" dropped               |
| `TYPE OF WORK`   | Employment classification                                                                        | Categorical (All employees / Full-time / Part-time)| Filtered to "All employees" for EDA; full-time-only is a useful sensitivity filter                                  |
| `WAGE RATE`      | Descriptor of the wage measure recorded in the file                                              | Categorical — single value: "Median hourly wage"  | Not used analytically; only one value exists in the dataset                                                         |
| `EDUCATION`      | Highest level of schooling attained by the worker group                                          | Ordered factor (8 levels)                          | Leading whitespace trimmed; ordered from lowest (0–8 years) to highest (Above bachelor's degree); "Total" rows dropped |
| `AGE GROUP`      | Age cohort of the worker group                                                                   | Categorical (15+, 25+, 25–34, 25–54, 25–64)       | Filtered to "25+" for main analysis to focus on prime working-age adults                                            |
| `Both sexes`     | Median hourly wage (CAD/hr) for both sexes combined — **primary outcome variable**               | Numeric (CAD/hour)                                 | No missing values in Ontario subset                                                                                  |
| `Men`            | Median hourly wage (CAD/hr) for men                                                              | Numeric (CAD/hour)                                 | Retained for gender-disaggregated analysis in later sections                                                        |
| `Women`          | Median hourly wage (CAD/hr) for women                                                            | Numeric (CAD/hour)                                 | Retained for gender-disaggregated analysis in later sections                                                        |


## Data Cleaning

Rows where `EDUCATION` or `IMMIGRANT` contained the word `"Total"` were removed prior to analysis to prevent double-counting, as these aggregate rows represent summaries across sub-groups already present in the data. The dataset was further filtered to Ontario only, since Canada-level rows would conflate Ontario's labour market with national trends outside this project's scope.

```{r data-clean-moiz, eval=TRUE}
library(tidyverse)

# Load the raw dataset
wages_raw <- readr::read_csv(
  "data.csv",
  trim_ws        = TRUE,
  show_col_types = FALSE
)

wages <- wages_raw %>%
  # Standardize column names to snake_case
  rename(
    year             = YEAR,
    geography        = GEOGRAPHY,
    immigrant_status = IMMIGRANT,
    work_type        = `TYPE OF WORK`,
    wage_type        = `WAGE RATE`,
    education_level  = EDUCATION,
    age_group        = `AGE GROUP`,
    wage_both        = `Both sexes`,
    wage_men         = Men,
    wage_women       = Women
  ) %>%
  # Focus on Ontario only (drop Canada-level aggregate rows)
  filter(str_trim(geography) == "Ontario") %>%
  filter(
    !str_detect(education_level, regex("total", ignore_case = TRUE)),
    !str_detect(immigrant_status, regex("total", ignore_case = TRUE))
  ) %>%
  mutate(across(where(is.character), str_trim)) %>%
  # Reorder education_level as an ordered factor from lowest to highest attainment
  mutate(
    education_level = factor(
      education_level,
      levels = c(
        "0 - 8 years",
        "Some high school",
        "High school graduate",
        "Some post-secondary",
        "Post-secondary certificate or diploma",
        "University degree",
        "Bachelor's degree",
        "Above bachelor's degree"
      ),
      ordered = TRUE
    )
  ) %>%
  filter(age_group == "25 +", work_type == "All employees")

glimpse(wages)
```

---

## Exploratory Data Analysis

The table below presents the mean, median, and standard deviation of the median hourly wage (`wage_both`, both sexes) for each education level, pooled across all years and immigrant status categories in the Ontario, age 25+ subset.

```{r table1-summary, eval=TRUE}
wages %>%
  group_by(education_level) %>%
  summarise(
    n           = n(),
    mean_wage   = mean(wage_both, na.rm = TRUE),
    median_wage = median(wage_both, na.rm = TRUE),
    sd_wage     = sd(wage_both, na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  knitr::kable(
    digits = 2,
    col.names = c(
      "Education Level", "N (obs.)",
      "Mean Hourly Wage ($)", "Median Hourly Wage ($)", "Std. Dev. ($)"
    ),
    format = "latex"
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width    = FALSE,
    font_size     = 10
  )
```

Table 1 reveals a clear positive wage gradient: mean hourly wages increase at each successive level of educational attainment, with university-level workers earning substantially more than those with eight or fewer years of schooling.

```{r graph1-immigrant-trend, eval=TRUE, fig.width=6, fig.height=4, fig.align='center', out.width='75%'}
wages_immig <- wages %>%
  mutate(
    immig_group = case_when(
      str_detect(immigrant_status, regex("born in canada", ignore_case = TRUE)) ~ "Born in Canada",
      str_detect(immigrant_status, regex("landed", ignore_case = TRUE))         ~ "Total Landed Immigrants",
      TRUE                                                                       ~ NA_character_
    )
  ) %>%
  filter(!is.na(immig_group)) %>%
  group_by(year, immig_group) %>%
  summarise(mean_wage = mean(wage_both, na.rm = TRUE), .groups = "drop")

ggplot(wages_immig, aes(x = year, y = mean_wage, 
                        colour = immig_group, 
                        group = immig_group)) +
  geom_line(linewidth = 1) + 
  geom_point(size = 1.8, alpha = 0.9) + 
  scale_colour_manual(
    values = c("Born in Canada" = "#2a9d8f", "Total Landed Immigrants" = "#f4a261"),
    name   = "Worker Group"
  ) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$", suffix = "/hr")) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 7)) +
  labs(
    title    = "Wage Trends Over Time: Immigrants vs. Canadian-Born Workers",
    subtitle = "Mean of median hourly wages (CAD/hr) — Ontario, age 25+",
    x        = "Year",
    y        = "Mean Wage (CAD/hr)",
    caption  = "Source: Ontario MLITSD. 'Total Landed Immigrants' includes all arrival categories."
  ) +
  theme_minimal(base_size = 9) + 
  theme(
    axis.title.y = element_text(margin = margin(r = 10)),
    plot.title = element_text(face = "bold", size = 10, hjust = 0.5), 
    plot.subtitle = element_text(colour = "grey40", size = 8, hjust = 0.5),
    plot.caption = element_text(colour = "grey50", size = 7, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )
```

**Interpretation.** Graph 1 compares the median hourly wage trajectories of landed immigrants and Canadian-born workers across all education levels in Ontario from 2006 to 2024. Two patterns stand out. First, a persistent wage gap is visible throughout the entire observation period: Canadian-born workers consistently earn more per hour than landed immigrants, even before controlling for education. Second, while both groups experience nominal wage growth over time, the gap does not meaningfully close, and in some periods appears to widen, suggesting that general labour market growth has not translated into convergence between the two groups. This is significant for our research question because it establishes that the immigrant wage disadvantage is not simply a credential issue: even when all education levels are pooled together, immigrants earn less.

```{r graph2-university-immigrant, eval=TRUE, fig.width=6, fig.height=4, fig.align='center', out.width='75%'}
wages_univ <- wages %>%
  filter(education_level == "University degree") %>%
  mutate(
    immig_group = case_when(
      str_detect(immigrant_status, regex("born in canada", ignore_case = TRUE)) ~ "Born in Canada",
      str_detect(immigrant_status, regex("landed", ignore_case = TRUE))         ~ "Total Landed Immigrants",
      TRUE                                                                       ~ NA_character_
    )
  ) %>%
  filter(!is.na(immig_group)) %>%
  group_by(year, immig_group) %>%
  summarise(mean_wage = mean(wage_both, na.rm = TRUE), .groups = "drop")

ggplot(wages_univ, aes(x = year, y = mean_wage, 
                        colour = immig_group, 
                        group = immig_group)) +
  geom_line(linewidth = 1) + geom_point(size = 1.8, alpha = 0.9) +      
  scale_colour_manual(
    values = c("Born in Canada" = "#2a9d8f", "Total Landed Immigrants" = "#f4a261"),
    name   = "Worker Group"
  ) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$", suffix = "/hr")) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 7)) + 
  labs(
    title    = "University Degree Holders: Immigrants vs. Canadian-Born",
    subtitle = "Mean of median hourly wages (CAD/hr) — Ontario, age 25+",
    x        = "Year",
    y        = "Mean Wage (CAD/hr)",           
    caption  = "Source: Ontario MLITSD. Data filtered to 'University degree' only."
  ) + theme_minimal(base_size = 9) +  theme(
    axis.title.y = element_text(margin = margin(r = 10)),
    plot.title = element_text(face = "bold", size = 10, hjust = 0.5),
    plot.subtitle = element_text(colour = "grey40", size = 8, hjust = 0.5),
    plot.caption = element_text(colour = "grey50", size = 7, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )
```

**Interpretation.** Graph 2 provides the most direct visual evidence relevant to the glass ceiling hypothesis examined in this report. By restricting the comparison to workers who hold a university degree, it controls for educational attainment and isolates the wage gap attributable to immigrant status alone. The persistent separation between the two lines with Canadian-born university graduates earning more than their immigrant counterparts in nearly every year from 2006 to 2024 suggests that a university credential does not yield equal labour market returns for all who hold it. Notably, the gap does not consistently narrow over time, which would be expected if the disadvantage were driven purely by short-term integration barriers such as language or unfamiliarity with Canadian workplace norms. Instead, the sustained gap points toward more structural explanations: non-recognition of foreign credentials, employer discrimination, or network disadvantages that persist even among established immigrants. This pattern directly motivates the formal hypothesis tests and regression model in the sections that follow.

\newpage

# Bootstrapping and Hypothesis Testing

## Hypothesis testing

We are testing whether mean wages differ by education level. For inference, we use a two-sample t-test on "University degree" vs "High school graduate" to estimate both significance and the 95% confidence interval for the mean difference.
Null hypothesis: The mean wage is the same for these two education groups.
Alternative hypothesis: The mean wage is different between these two education groups.

```{r hypothesis-tests-education-asad, eval=TRUE}
wages_without_total <- wages %>% filter(education_level != "Total, all education levels")

# Summaries use the numeric wage columns (WAGE RATE is only the label "Median hourly wage").
wages_by_education <- wages_without_total %>%
  group_by(education_level) %>%
  summarise(mean_wage = mean(wage_both, na.rm = TRUE), .groups = "drop")
kable(wages_by_education, caption = "Mean of wage_both (median hourly wage) by education level")

# t-test needs exactly TWO groups on the right-hand side — filter to two education levels, then test on row-level values.
two_edu <- wages_without_total %>%
  filter(education_level %in% c("University degree", "High school graduate"))
edu_test <- t.test(wage_both ~ education_level, data = two_edu)
kable(
  ttest_to_long_table(
    edu_test,
    "High school graduate",
    "University degree",
    "High school graduate - University degree"
  ),
  caption = "Welch two-sample t-test for high school graduates vs university degree holders"
)
```
**Interpretation.** The summary table compares all education categories, while the t-test focuses on "University degree" versus "High school graduate." The 95% confidence interval for the mean difference (High school - University) is approximately [-13.07, -10.42], which does not include 0. Therefore, we reject the null hypothesis and conclude with at least 95% confidence that university-degree workers have a higher mean wage than high-school graduates. This directly supports the research question that education level is associated with wage differences.

Next, we test whether mean wages differ by immigrant status using a two-sample t-test for "Born in Canada" versus "Established immigrants, 10+ years," again using the test to obtain a 95% confidence interval for the mean difference.
Null hypothesis: The mean wage is the same for these two immigrant-status groups.
Alternative hypothesis: The mean wage is different between these two immigrant-status groups.
```{r hypothesis-tests-immigrant-asad, eval=TRUE}
wages_without_total <- wages %>% filter(immigrant_status != "Total, all immigrant status")

# Summaries use the numeric wage columns (WAGE RATE is only the label "Median hourly wage").
wages_by_immigrant_status <- wages_without_total %>%
  group_by(immigrant_status) %>%
  summarise(mean_wage = mean(wage_both, na.rm = TRUE), .groups = "drop")
kable(wages_by_immigrant_status, caption = "Mean of wage_both (median hourly wage) by immigrant status")

two_immigrant_status <- wages_without_total %>%
  filter(immigrant_status %in% c("Born in Canada", "Established immigrants, 10+ years"))
immigrant_test <- t.test(wage_both ~ immigrant_status, data = two_immigrant_status)
kable(
  ttest_to_long_table(
    immigrant_test,
    "Born in Canada",
    "Established immigrants, 10+ years",
    "Born in Canada - Established immigrants, 10+ years"
  ),
  caption = "Welch two-sample t-test for Canadian-born vs established immigrants"
)
```

**Interpretation.** The summary table shows all immigrant-status categories, and the t-test isolates "Born in Canada" versus "Established immigrants, 10+ years." The 95% confidence interval for the mean difference (Born in Canada - Established immigrants) is approximately [0.97, 4.82], which does not include 0. Therefore, we reject the null hypothesis and conclude with at least 95% confidence that Canadian-born workers in this comparison have a higher mean wage. This supports the second research question by showing a wage gap by immigrant status(the highest wage average immigrant status in fact).

Finally, we test whether immigrant-status wage differences remain within the university-educated subgroup using a two-sample t-test for "Born in Canada" versus "Established immigrants, 10+ years," and we use its 95% confidence interval for the mean difference.
Null hypothesis: The mean wage is the same for these two immigrant-status groups among university-educated workers.
Alternative hypothesis: The mean wage is different between these two immigrant-status groups among university-educated workers.
```{r hypothesis-tests-university-immigrant-asad, eval=TRUE}
wages_without_total <- wages %>% filter(education_level == "University degree")
wages_by_immigrant_status <- wages_without_total %>%
  group_by(immigrant_status) %>%
  summarise(mean_wage = mean(wage_both, na.rm = TRUE), .groups = "drop")
kable(wages_by_immigrant_status, caption = "Mean of wage_both (median hourly wage) by immigrant status")

two_immigrant_status <- wages_without_total %>%
  filter(immigrant_status %in% c("Born in Canada", "Established immigrants, 10+ years"))
university_immigrant_test <- t.test(wage_both ~ immigrant_status, data = two_immigrant_status)
kable(
  ttest_to_long_table(
    university_immigrant_test,
    "Born in Canada",
    "Established immigrants, 10+ years",
    "Born in Canada - Established immigrants, 10+ years"
  ),
  caption = "Welch two-sample t-test within university graduates"
)
```

**Interpretation.** This final t-test focuses only on university-educated respondents. The 95% confidence interval for the mean difference (Born in Canada - Established immigrants) is approximately [1.88, 7.13], which does not include 0. Therefore, we reject the null hypothesis and conclude with at least 95% confidence that, even among university graduates, Canadian-born workers in this comparison have a higher mean wage. This addresses the third research question and supports the interpretation that the gap is not only an education effect.

Taken together, these three t-tests show that wages differ significantly by education level, by immigrant status, and within the university-educated subgroup by immigrant status. Because none of the estimated confidence intervals include 0, each comparison provides evidence of a meaningful difference in mean wages. Overall, this strengthens the report's central conclusion: education is associated with higher wages, but the returns to education do not appear to be distributed equally across immigrant groups, which is consistent with the glass-ceiling framing in Ontario's labour market.

## Bootstrapping

Bootstrapping gives us a second way to assess uncertainty in the estimated wage gaps. Unlike the t-test, which relies on distributional assumptions, bootstrap resampling repeatedly samples from the observed data with replacement and builds an empirical distribution of the mean difference. This is especially useful here because wage data can be skewed and may not follow a perfectly normal distribution.

In this section, we bootstrap the same three comparisons used in the hypothesis tests: university graduates versus high school graduates, Canadian-born workers versus established immigrants, and Canadian-born versus established immigrants within the university-educated subgroup. For each comparison, we use 1000 bootstrap resamples and report the 95% bootstrap confidence interval for the difference in mean wages.

```{r bootstrap-asad, fig.cap="Bootstrap distributions for the mean wage differences across the three main comparisons.", eval=TRUE, fig.width=7.5, fig.height=9, fig.align='center', out.width='95%'}
set.seed(123)

bootstrap_mean_diff <- function(group_a, group_b, reps = 1000) {
  group_a <- group_a[!is.na(group_a)]
  group_b <- group_b[!is.na(group_b)]
  replicate(reps, {
    mean(sample(group_a, size = length(group_a), replace = TRUE)) -
      mean(sample(group_b, size = length(group_b), replace = TRUE))
  })
}

university <- wages %>% filter(education_level == "University degree")
highschool <- wages %>% filter(education_level == "High school graduate")
education_boot <- bootstrap_mean_diff(university$wage_both, highschool$wage_both)

born_canada <- wages %>% filter(immigrant_status == "Born in Canada")
established_immigrants <- wages %>% filter(immigrant_status == "Established immigrants, 10+ years")
immigrant_boot <- bootstrap_mean_diff(born_canada$wage_both, established_immigrants$wage_both)

university_only <- wages %>% filter(education_level == "University degree")
born_canada_uni <- university_only %>% filter(immigrant_status == "Born in Canada")
established_immigrants_uni <- university_only %>% filter(immigrant_status == "Established immigrants, 10+ years")
university_immigrant_boot <- bootstrap_mean_diff(
  born_canada_uni$wage_both,
  established_immigrants_uni$wage_both
)

bootstrap_plot_data <- tibble(
  mean_difference = c(
    education_boot,
    immigrant_boot,
    university_immigrant_boot
  ),
  comparison = factor(
    c(
      rep("University degree - High school graduate", length(education_boot)),
      rep("Born in Canada - Established immigrants, 10+ years", length(immigrant_boot)),
      rep("Born in Canada - Established immigrants, 10+ years\n(University only)", length(university_immigrant_boot))
    ),
    levels = c(
      "University degree - High school graduate",
      "Born in Canada - Established immigrants, 10+ years",
      "Born in Canada - Established immigrants, 10+ years\n(University only)"
    )
  )
)

bootstrap_ci_lines <- tibble(
  comparison = factor(
    c(
      "University degree - High school graduate",
      "University degree - High school graduate",
      "Born in Canada - Established immigrants, 10+ years",
      "Born in Canada - Established immigrants, 10+ years",
      "Born in Canada - Established immigrants, 10+ years\n(University only)",
      "Born in Canada - Established immigrants, 10+ years\n(University only)"
    ),
    levels = levels(bootstrap_plot_data$comparison)
  ),
  xintercept = c(
    quantile(education_boot, probs = c(0.025, 0.975)),
    quantile(immigrant_boot, probs = c(0.025, 0.975)),
    quantile(university_immigrant_boot, probs = c(0.025, 0.975))
  )
)

ggplot(bootstrap_plot_data, aes(x = mean_difference)) +
  geom_histogram(bins = 30, fill = "#8fb9a8", color = "white") +
  geom_vline(
    data = bootstrap_ci_lines,
    aes(xintercept = xintercept),
    color = "#c0392b",
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  facet_wrap(~comparison, ncol = 1, scales = "free_y") +
  labs(
    title = "Bootstrap Distributions of Mean Wage Differences",
    x = "Bootstrap mean difference",
    y = "Count"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

bootstrap_summary <- tibble(
  comparison = c(
    "University degree - High school graduate",
    "Born in Canada - Established immigrants, 10+ years",
    "Born in Canada - Established immigrants, 10+ years (University only)"
  ),
  bootstrap_mean_difference = c(
    mean(education_boot),
    mean(immigrant_boot),
    mean(university_immigrant_boot)
  ),
  ci_lower = c(
    quantile(education_boot, 0.025),
    quantile(immigrant_boot, 0.025),
    quantile(university_immigrant_boot, 0.025)
  ),
  ci_upper = c(
    quantile(education_boot, 0.975),
    quantile(immigrant_boot, 0.975),
    quantile(university_immigrant_boot, 0.975)
  )
)

bootstrap_summary_long <- tibble(
  Comparison = rep(bootstrap_summary$comparison, each = 3),
  Metric = rep(c("Bootstrap mean difference", "95% CI lower", "95% CI upper"), times = nrow(bootstrap_summary)),
  Value = c(
    bootstrap_summary$bootstrap_mean_difference[1],
    bootstrap_summary$ci_lower[1],
    bootstrap_summary$ci_upper[1],
    bootstrap_summary$bootstrap_mean_difference[2],
    bootstrap_summary$ci_lower[2],
    bootstrap_summary$ci_upper[2],
    bootstrap_summary$bootstrap_mean_difference[3],
    bootstrap_summary$ci_lower[3],
    bootstrap_summary$ci_upper[3]
  )
)

kable(
  bootstrap_summary_long,
  digits = 3,
  caption = "Bootstrap 95% confidence intervals for mean wage differences"
)
```

**Interpretation.** The bootstrap results reinforce the conclusions from the t-tests. If the 95% bootstrap confidence interval for a mean difference does not include 0, that indicates the estimated wage gap remains consistently positive or negative across repeated resamples of the data. For the education comparison, this supports the conclusion that university graduates earn more on average than high school graduates. For the two immigrant-status comparisons, bootstrap intervals that remain above 0 support the conclusion that Canadian-born workers earn more on average than established immigrants, both overall and within the university-educated subgroup.

Overall, the bootstrap analysis strengthens the inference section by showing that the main findings are not driven only by the assumptions of the t-test. Because the resampled confidence intervals tell the same story as the analytical confidence intervals, we have stronger evidence that the observed wage gaps are real and consistent with the report's glass-ceiling interpretation.

\newpage

# The Model and Conclusion

## Regression analysis
```{r regression-ishan, eval=TRUE}
# 1. Prepare data with explicit reference levels
wages_model <- wages %>%
  mutate(
    education_level = relevel(factor(as.character(education_level), ordered = FALSE), 
                              ref = "High school graduate"),
    immigrant_status = relevel(factor(immigrant_status), 
                               ref = "Born in Canada")
  ) %>%
  filter(
    !is.na(wage_both),
    !is.na(education_level),
    !is.na(immigrant_status)
  )

wages_model <- wages_model %>%
  mutate(year_centered = year - 2006)

fit_mlr <- lm(wage_both ~ education_level + immigrant_status + year_centered, data = wages_model)

# 3. Extract and format results
coef_mat <- summary(fit_mlr)$coefficients
ci_mat <- confint(fit_mlr)

reg_table <- tibble::tibble(
  term = rownames(coef_mat),
  estimate = coef_mat[, 1],
  std_error = coef_mat[, 2],
  p_value = coef_mat[, 4],
  ci_lower = ci_mat[, 1],
  ci_upper = ci_mat[, 2]
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

reg_table_clean <- reg_table %>%
  mutate(term = stringr::str_remove_all(term, "education_level|immigrant_status")) %>%
  mutate(term = ifelse(term == "(Intercept)", "Baseline (High School / Born in Canada)", term))

important_terms <- c(
  "Baseline (High School / Born in Canada)",
  "0 - 8 years",
  "Some post-secondary",
  "Bachelor's degree",
  "University degree",
  "Above bachelor's degree",
  "Established immigrants, 10+ years",
  "Recent immigrants, 5+ to 10 years",
  "Very recent immigrants, 5 years or less",
  "year_centered"
)

reg_table_compact <- reg_table_clean %>%
  filter(term %in% important_terms) %>%
  mutate(
    term = dplyr::case_when(
      term == "Baseline (High School / Born in Canada)" ~ "Reference: High school graduate + Born in Canada (year 2006)",
      term %in% c("0 - 8 years", "Some post-secondary", "Bachelor's degree", "University degree", "Above bachelor's degree") ~
        paste0("Education: ", term, " (vs High school graduate)"),
      term %in% c("Established immigrants, 10+ years", "Recent immigrants, 5+ to 10 years", "Very recent immigrants, 5 years or less") ~
        paste0("Immigrant status: ", term, " (vs Born in Canada)"),
      term == "year_centered" ~ "Year centered at 2006 (per +1 year)",
      TRUE ~ term
    ),
    p_value = ifelse(p_value < 0.001, "<0.001", sprintf("%.3f", p_value))
  ) %>%
  transmute(
    Variable = term,
    Estimate = round(estimate, 3),
    `p-value` = p_value
  )

knitr::kable(
  reg_table_compact,
  booktabs = TRUE,
  caption = "Key Regression Coefficients for Median Hourly Wage (CAD/hr)"
)
```
**Interpretation.** The linear regression model shows that education is strongly associated with higher wages, but the wage gap for those with immigrant-status remains. The baseline group (high school graduates born in Canada in 2006) has a predicted wage of 18.595 CAD/hr. Compared to this baseline, those with university-level education have higher wages (University degree +11.745, Bachelor’s +9.196, Above bachelor’s +16.303 CAD/hr, all p < 0.001 so the results are strongly supported). Those with lower education categories are associated with lower wages. The only education category without a clear difference is some post-secondary (-0.147 CAD/hr, p = 0.744, but as p value is high, results are not as accurate).

Those with an immigrant-status earn less than Canadian-born workers, even after controlling education and year (established immigrants -2.895, recent immigrants 5+ to 10 years -7.954, very recent immigrants -10.259 CAD/hr, all p < 0.001 so the results are strongly supported). The positive year effect (+0.508 CAD/hr, p < 0.001) indicates general wage growth over time but not nearly as high as Canadian-born workers. Overall, the model supports the main conclusion that education is rewarded, but the return is not equally distributed across immigrant groups indicating the presence of a glass-ceiling.


## Cross-validation
```{r cross-validation-ishan, eval=TRUE}
# 4-fold cross-validation
set.seed(123)

d <- wages %>%
  mutate(
    education_level = relevel(factor(as.character(education_level), ordered = FALSE),
      ref = "High school graduate"
    ),
    immigrant_status = relevel(factor(immigrant_status), ref = "Born in Canada"),
    year_centered = year - 2006
  ) %>%
  filter(!is.na(wage_both), !is.na(education_level), 
         !is.na(immigrant_status), !is.na(year_centered))

k <- 4
d <- d %>% mutate(group_ind = sample(1:k, size = nrow(d), replace = TRUE))
mse_vec <- numeric(k)

for (i in 1:k) {
  d.train <- d %>% filter(group_ind != i)
  d.test <- d %>% filter(group_ind == i)
  lm.mod <- lm(wage_both ~ education_level + immigrant_status + year_centered, data = d.train)
  y_hat <- predict(lm.mod, newdata = d.test)
  mse_vec[i] <- mean((d.test$wage_both - y_hat)^2, na.rm = TRUE)
}

# --- TABLE 1: Detailed Fold Results ---
cv_results <- tibble::tibble(
  Fold = paste0("Fold ", 1:k),
  `MSE ((CAD/hr)^2)` = round(mse_vec, 3)
)

knitr::kable(
  cv_results, 
  booktabs = TRUE, 
  caption = "4-Fold Cross-Validation: Mean Squared Error by Fold",
  escape = FALSE # Allows the superscript in the header to render correctly
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")

# --- TABLE 2: Summary Metrics ---
overall_mse <- mean(mse_vec)
overall_rmse <- sqrt(overall_mse)

summary_metrics <- tibble::tibble(
  Metric = c("Average MSE", "Average RMSE"),
  Unit = c("(CAD/hr)^2", "CAD/hr"),
  Value = round(c(overall_mse, overall_rmse), 3)
)

knitr::kable(
  summary_metrics, 
  booktabs = TRUE, 
  caption = "Overall Model Predictive Performance",
  col.names = c("Metric", "Unit", "Value")
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```
**Interpretation.** We evaluated the regression model performance using a 4-fold cross-validation. In each fold, the model was trained on 3 parts of the data and tested on the remaining part, then MSE was computed on the test fold. MSE values ranged from 9.105 to 14.291, with an average of 11.649. The corresponding average RMSE is 3.413 CAD/hr, meaning predictions are typically off by about $3.4 per hour on unseen data. Overall, the model shows a fairly stable predictive performance across folds.

## Summary and conclusion

Overall, the evidence indicates that education is strongly associated with higher wages in Ontario, but these returns are not equally distributed across immigrant groups. Analysis shows a clear wage difference by education level and an increasing education-related wage gap over time. Inference results and bootstrap confidence intervals consistently show significant differences in wages by both education and immigrant status, including within the university-educated subgroup.

The regression model confirms that higher education categories are associated with higher hourly wages after controlling for time, while several immigrant-status groups remain at a wage disadvantage relative to Canadian-born workers. Cross-validation results also suggest moderate and stable out-of-sample predictive performance, supporting the reliability of the model findings.

Taken together, these results support the glass-ceiling interpretation that education does improve earnings on average, but these earnings are not equally distributed across groups. This suggests the need for stronger educational recognition and improved newcomer wages, to ensure fairer wage returns to education in Ontario regardless of status.

\newpage

# Appendix: All code

**LABEL (remove before submission):** Ensure this lists the final, clean versions of everyone's chunks (knitr will pull in the labeled chunks below). If you rename chunks, update `ref.label` accordingly.

```{r appendix-all-code, ref.label=c("libraries", "data-load", "table1-summary", "graph1-immigrant-trend", "graph2-university-immigrant", "hypothesis-tests-education-asad", "hypothesis-tests-immigrant-asad", "hypothesis-tests-university-immigrant-asad", "ci-asad", "bootstrap-asad", "regression-ishan", "cross-validation-ishan"), echo=TRUE, eval=FALSE}
```

# References

Ministry of Labour, Immigration, training and Skills Development. ontario.ca. (2025, August 12). https://www.ontario.ca/page/ministry-labour-immigration-training-skills-development 

Ontario.ca. (n.d.). Wages by education level. Wages by education level - Dataset - Ontario Data Catalogue. https://data.ontario.ca/dataset/wages-by-education-level 