---
title: "The Glass Ceiling: Does Education Equitably Predict Wages for Ontario Immigrants?"
author:
  - "Asad Mirza"
  - "Moiz Mohammad"
  - "Ishan Amin"
date: "`r Sys.Date()`"
documentclass: article
fontsize: 11pt
geometry: margin=1in
header-includes:
  - '\linespread{1.15}\selectfont'
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Project overview

This project examines whether **education** predicts **wages** in a similar way for **immigrants in Ontario** as it might for other groups—or whether systematic differences suggest barriers consistent with a **glass ceiling** (e.g., credentials not translating equally into earnings). We use data-science methods to describe the relationship between schooling and income and to test whether that relationship differs in ways that raise equity questions.

## Core team

| Role        | Name          |
|-------------|---------------|
| Core worker | Asad Mirza    |
| Core worker | Moiz Mohammad |
| Core worker | Ishan Amin    |

## Research questions (draft)

1. How strongly is educational attainment associated with wages among Ontario residents with an immigrant background?
2. Does the education–wage gradient differ by gender, region within Ontario, immigrant class, or years since landing (hypotheses tied to “glass ceiling” and unequal returns to human capital)?
3. After controlling for relevant covariates, do we still see disparities that suggest education does **not** equitably predict wages for this population?

## Data and methods (to be specified)

- **Data sources:** To be finalized (e.g., public microdata or summary tables from Statistics Canada relevant to labour income, education, immigration status, and geography—Ontario).
- **Variables:** Outcome (e.g., wages or earnings); key predictor (education); controls (age, gender, tenure, occupation/industry if available, etc.).
- **Analysis:** Exploratory summaries and visualizations; regression or similar models to quantify associations and interactions; sensitivity checks as appropriate.

```{r libraries}
# Add packages as the analysis develops, e.g.:
# library(tidyverse)
```

## Next steps

- Lock in data source(s) and access path (file paths or API keys in a local `.Renviron`, not committed).
- Reproduce a clean analytic sample (Ontario, immigrants, key variables complete).
- Draft results and interpretation with explicit connection to the “glass ceiling” framing.

## References

- Placeholder: cite Statistics Canada products and any methods papers once chosen.
